---
title: "What Drove Crop Production Growth in Pakistan and India: Yield Improvement or Land Expansion?"
author: "Shrish Tariq"
format: html
execute: 
  echo: false
---

```{r}
library(tidyverse)
install.packages("ggplot2")
library(ggplot2)
library(scales)
```

Loading agriculture dataset
```{r}
agri_data <- read.csv(
  "data/agriculture.csv",
  fileEncoding = "UTF-8-BOM"
)
```

Exploring agriculture dataset

```{r}
library(dplyr)
library(tidyr)

crop_data <- agri_data %>%
  filter(
    Area %in% c("Pakistan", "India"),
    Item %in% c("Wheat", "Rice", "Maize (corn)"),
    Element %in% c("Area harvested", "Yield", "Production"),
    !is.na(Value)
  ) %>%
  mutate(
    Item = recode(Item, "Maize (corn)" = "Maize")
  ) %>%
  select(Area, Item, Year, Element, Value) %>%
  pivot_wider(
    names_from = Element,
    values_from = Value
  ) %>%
  rename(
    country = Area,
    crop = Item,
    year = Year,
    area_ha = `Area harvested`,
    yield_kg_ha = Yield,
    production_t = Production
  ) %>%
  arrange(country, crop, year)
```
```{r}
head(agri_data)
str(agri_data)
dim(agri_data)
names(agri_data)
```

```{r}
head(crop_data)
dim(crop_data)

crop_data %>%
  count(country, crop)
```

How have wheat, rice, and maize production trends differed between Pakistan and India since 2011?
```{r}
library(ggplot2)
library(scales)

ggplot(
  crop_data,
  aes(x = year, y = production_t, color = country)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ crop, scales = "free_y") +
  scale_y_continuous(
    labels = label_number(scale = 1e-6, suffix = "M")
  ) +
  labs(
    title = "Crop Production Trends in Pakistan and India",
    subtitle = "Wheat, rice and maize production, 2011–2023",
    x = "Year",
    y = "Production (million tonnes)",
    color = "Country",
    caption = "Source: FAOSTAT"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Which crop-country combination experienced the fastest production growth?

```{r}
library(dplyr)
library(ggplot2)
indexed_data <- crop_data %>% group_by(country, crop) %>% mutate(production_2011 = production_t[year == 2011][1], production_index = (production_t / production_2011) * 100
) %>%
    ungroup()
indexed_data %>%
  filter(year == 2011) %>%
  select(country, crop, year, production_index)
```

```{r}
ggplot(
  indexed_data,
  aes(
    x = year,
    y = production_index,
    color = country,
    linetype = country,
    group = country
  )
) +
  geom_hline(
    yintercept = 100,
    color = "grey50",
    linetype = "dotted"
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ crop, nrow = 1) +
  scale_color_manual(
    values = c(
      "India" = "#2563EB",
      "Pakistan" = "#D97706"
    )
  ) +
  scale_x_continuous(
    breaks = seq(2011, 2023, by = 2)
  ) +
  labs(
    title = "Indexed Crop Production in Pakistan and India",
    subtitle = "Each crop–country series is set to 100 in 2011",
    x = "Year",
    y = "Production index (2011 = 100)",
    color = "Country",
    linetype = "Country",
    caption = "Source: FAOSTAT"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

Growth ranking of Pakistan and India in terms of economic crops
```{r}
growth_ranking <- indexed_data %>%
  group_by(country, crop) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    country,
    crop,
    final_year = year,
    final_index = round(production_index, 1),
    growth_percent = round(production_index - 100, 1)
  ) %>%
  arrange(desc(growth_percent))

growth_ranking
```

Was production growth driven by better yields or expansion of cultivated area?

```{r}
growth_summary <- crop_data %>%
  group_by(country, crop) %>%
  summarise(
    area_change_pct = 100 * (
      last(area_ha, order_by = year) /
        first(area_ha, order_by = year) - 1
    ),
    yield_change_pct = 100 * (
      last(yield_kg_ha, order_by = year) /
        first(yield_kg_ha, order_by = year) - 1
    ),
    production_change_pct = 100 * (
      last(production_t, order_by = year) /
        first(production_t, order_by = year) - 1
    ),
    .groups = "drop"
  )

print(growth_summary)
```