---
title: "What Drove Crop Production Growth in Pakistan and India: Yield Improvement or Land Expansion?"
author: "Shrish Tariq"
format: html
execute: 
  echo: false
---

```{r}
library(tidyverse)
install.packages("ggplot2")
library(ggplot2)
library(scales)
```

Loading agriculture dataset
```{r}
agri_data <- read.csv(
  "data/agriculture.csv",
  fileEncoding = "UTF-8-BOM"
)
```

Exploring agriculture dataset

```{r}
library(dplyr)
library(tidyr)

crop_data <- agri_data %>%
  filter(
    Area %in% c("Pakistan", "India"),
    Item %in% c("Wheat", "Rice", "Maize (corn)"),
    Element %in% c("Area harvested", "Yield", "Production"),
    !is.na(Value)
  ) %>%
  mutate(
    Item = recode(Item, "Maize (corn)" = "Maize")
  ) %>%
  select(Area, Item, Year, Element, Value) %>%
  pivot_wider(
    names_from = Element,
    values_from = Value
  ) %>%
  rename(
    country = Area,
    crop = Item,
    year = Year,
    area_ha = `Area harvested`,
    yield_kg_ha = Yield,
    production_t = Production
  ) %>%
  arrange(country, crop, year)
```
```{r}
head(agri_data)
str(agri_data)
dim(agri_data)
names(agri_data)
```

```{r}
head(crop_data)
dim(crop_data)

crop_data %>%
  count(country, crop)
```

How have wheat, rice, and maize production trends differed between Pakistan and India since 2011?
```{r}
library(ggplot2)
library(scales)

ggplot(
  crop_data,
  aes(x = year, y = production_t, color = country)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ crop, scales = "free_y") +
  scale_y_continuous(
    labels = label_number(scale = 1e-6, suffix = "M")
  ) +
  labs(
    title = "Crop Production Trends in Pakistan and India",
    subtitle = "Wheat, rice and maize production, 2011–2023",
    x = "Year",
    y = "Production (million tonnes)",
    color = "Country",
    caption = "Source: FAOSTAT"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Which crop-country combination experienced the fastest production growth?

```{r}
library(dplyr)
library(ggplot2)
indexed_data <- crop_data %>% group_by(country, crop) %>% mutate(production_2011 = production_t[year == 2011][1], production_index = (production_t / production_2011) * 100
) %>%
    ungroup()
indexed_data %>%
  filter(year == 2011) %>%
  select(country, crop, year, production_index)
```

```{r}
ggplot(
  indexed_data,
  aes(
    x = year,
    y = production_index,
    color = country,
    linetype = country,
    group = country
  )
) +
  geom_hline(
    yintercept = 100,
    color = "grey50",
    linetype = "dotted"
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ crop, nrow = 1) +
  scale_color_manual(
    values = c(
      "India" = "#2563EB",
      "Pakistan" = "#D97706"
    )
  ) +
  scale_x_continuous(
    breaks = seq(2011, 2023, by = 2)
  ) +
  labs(
    title = "Indexed Crop Production in Pakistan and India",
    subtitle = "Each crop–country series is set to 100 in 2011",
    x = "Year",
    y = "Production index (2011 = 100)",
    color = "Country",
    linetype = "Country",
    caption = "Source: FAOSTAT"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

Growth ranking of Pakistan and India in terms of economic crops
```{r}
growth_ranking <- indexed_data %>%
  group_by(country, crop) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    country,
    crop,
    final_year = year,
    final_index = round(production_index, 1),
    growth_percent = round(production_index - 100, 1)
  ) %>%
  arrange(desc(growth_percent))

growth_ranking
```

Was production growth driven by better yields or expansion of cultivated area?

```{r}
growth_summary <- crop_data %>%
  group_by(country, crop) %>%
  summarise(
    area_change_pct = 100 * (
      last(area_ha, order_by = year) /
        first(area_ha, order_by = year) - 1
    ),
    yield_change_pct = 100 * (
      last(yield_kg_ha, order_by = year) /
        first(yield_kg_ha, order_by = year) - 1
    ),
    production_change_pct = 100 * (
      last(production_t, order_by = year) /
        first(production_t, order_by = year) - 1
    ),
    .groups = "drop"
  )

print(growth_summary)
```

```{r}
growth_long <- growth_summary %>%
  pivot_longer(
    cols = ends_with("_change_pct"),
    names_to = "measure",
    values_to = "change_pct"
  ) %>%
  mutate(
    measure = recode(
      measure,
      area_change_pct = "Harvested area",
      yield_change_pct = "Yield",
      production_change_pct = "Production"
    ),
    measure = factor(
      measure,
      levels = c("Harvested area", "Yield", "Production")
    )
  )

p3 <- ggplot(
  growth_long,
  aes(x = crop, y = change_pct, fill = measure)
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_hline(yintercept = 0, colour = "grey40") +
  facet_wrap(~country) +
  scale_fill_manual(
    values = c(
      "Harvested area" = "#80B1D3",
      "Yield" = "#B3DE69",
      "Production" = "#FB8072"
    )
  ) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "What drove production growth?",
    subtitle = "Percentage change between 2011 and 2023",
    x = NULL,
    y = "Change (%)",
    fill = NULL,
    caption = "Source: FAOSTAT"
  )

p3

ggsave(
  "figures/03_growth_drivers.png",
  p3,
  width = 10,
  height = 6,
  dpi = 300
)
```

How have crop yields changed in the two countries?

```{r}
library(dplyr)
library(ggplot2)

country_colours <- c(
  "Pakistan" = "#D97706",
  "India" = "#2563EB"
)
p4 <- ggplot(
  crop_data,
  aes(x = year, y = yield_kg_ha, colour = country)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~crop, scales = "free_y") +
  scale_colour_manual(values = country_colours) +
  scale_x_continuous(breaks = seq(2011, 2023, 2)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Crop-yield trends in Pakistan and India",
    subtitle = "Yield represents output per hectare",
    x = NULL,
    y = "Yield (kg/ha)",
    colour = "Country",
    caption = "Source: FAOSTAT"
  )

p4

ggsave(
  "figures/04_yield_trends.png",
  p4,
  width = 11,
  height = 6,
  dpi = 300
)
```